---
title: "Exercise 02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--Original file: hw02_explore-gapminder-dplyr.md-->

# Explore Gapminder and use `dplyr`

## Overview

The goal is to explore a new-to-you dataset. In particular, to begin to establish a workflow for data frames or “tibbles”. You will use `dplyr` and `ggplot2` to do some description and visualization.

Remember the [sampler][wiki-sampler] concept. Your work should serve as your own personal cheatsheet in the future for things to do with a new dataset. Give yourself the cheatsheet you deserve!


## Bring rectangular data in

Work with the `gapminder` data. The Gapminder data is distributed as an [R package on CRAN][gapminder-cran].

Install `gapminder` if you have not done so already and remember to load it.

``` r
install.packages("gapminder")
library(gapminder)
```

Install and load `dplyr.` Probably via the `tidyverse` meta-package.

``` r
install.packages("tidyverse")
library(tidyverse)
```

## Smell test the data

Explore the `gapminder` object:

* Is it a data.frame, a matrix, a vector, a list?
* What's its class?
* How many variables/columns?
* How many rows/observations?
* Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?
* What flavor is each variable?
  
## Explore individual variables

Pick __at least__ one categorical variable and at least one quantitative variable to explore.

* Characterize what's possible, i.e. all possible values or max vs. min ... whatever's appropriate.
* What's typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.
* Feel free to use summary stats, tables, figures. We're NOT expecting high production value (yet).

## Explore various plot types

See the [`ggplot2` tutorial](https://github.com/jennybc/ggplot2-tutorial), which also uses the `gapminder` data, for ideas. 

Make a few plots, probably of the same variable you chose to characterize numerically. Try to explore more than one plot type. **Just as an example** of what I mean:

* A scatterplot of two quantitative variables.
* A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
* A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.

You don't have to use all the data in every plot! It's fine to filter down to one country or small handful of countries.

## Use `filter()`, `select()`, and `%>%`

## But I want to do more!

## Report your process

## Assess your work


```{r links, child="links.md"}
```