# Lab 06: Data wrangling conclusion {#lab-06}

```{r include = FALSE}
source("common.R")
```

<!--Original file: hw06_data-wrangling-conclusion.md-->

## Overview

Pick (at least) two of the topics below and do one of the exercise prompts listed, or something comparable.

## Character data

Read and work through the exercises in the [Strings chapter][r4ds-strings] of [R for Data Science][r4ds].


## Gapminder data

Work with the `gapminder` data. The Gapminder data is distributed as an [R package on CRAN][gapminder-cran].

Install `gapminder` if you have not done so already and remember to load it.

```{r eval = FALSE}
install.packages("gapminder")
library(gapminder)
```


## Writing functions

Pick one of the following:

### Option 1

Write one (or more) functions that do something useful to pieces of the Gapminder data. It is logical to think about computing on the mini-data frames
corresponding to the data for each specific country. This would pair well with the prompt below about working with a nested data frame, as you could apply your function there.

Make it something you can't easily do with built-in functions. Make it something that's not trivial to do with the simple `dplyr` verbs. The linear regression function [presented here](#functions-practicum) is a good starting point. You could generalize that to do quadratic regression (include a squared term) or use robust regression, using `MASS::rlm()` or `robustbase::lmrob()`.

### Option 2

If you plan to complete the lab where we build an R package, write a couple of experimental functions exploring some functionality that is useful to you in real life and that might form the basis of your personal package.
 
## Work with the candy data

## Work with a list

## Work with a nested data frame

## Report your process

## Assess your work



```{r links, child="links.md"}
```