# Lab 03: Use `dplyr` to manipulate and explore data (also use `ggplot2`) {#lab-03}

```{r include = FALSE}
source("common.R")
```

<!--Original file: hw03_dplyr-and-more-ggplot2.md-->
<!--APH adaptation: https://apreshill.github.io/data-vis-labs-2018/04-challenge.html-->

## Overview

The goal is to manipulate and explore a dataset with the `dplyr` package, complemented by visualizations made with `ggplot2`.

Remember the [sampler][wiki-sampler] concept. Your work should serve as your own personal cheatsheet in the future for ways to manipulate a dataset and produce companion figures. Give yourself the cheatsheet you deserve!

## Gapminder data

Work with the `gapminder` data. The Gapminder data is distributed as an [R package on CRAN][gapminder-cran].

Install `gapminder` if you have not done so already and remember to load it.

```{r eval = FALSE}
install.packages("gapminder")
library(gapminder)
```

## Your mission, high-level

Pick at least three of the tasks in the Task Menu section and attack each with a table and figure.

* `dplyr` should be your data manipulation tool
* `ggplot2` should be your visualization tool
  
Make observations about what your tables/figures show and about the process.

If you want to do something comparable but different, i.e. swap one quantitative variable for another, be my guest! If you are feeling inspired and curious, then we're doing this right. Go for it.

Relax about the following things:

* Tidying/reshaping is NOT your assignment. Many of your tables will be awkwardly shaped in the report. That's OK.
* Table beauty is not a big deal. Simply printing to "screen" is fine. You could also try the `knitr::kable()` function. Assuming `my_df` is a data.frame, here's an R chunk that should print it as a decent-looking table:

````
`r ''````{r results = 'asis'}
knitr::kable(my_df)
```
````

* For all things, graphical and tabular, if you're dissatisfied with a result, discuss the problem, what you've tried and move on.

## Task Menu

* Get the maximum and minimum of GDP per capita for all continents.

* Look at the spread of GDP per capita within the continents.

* Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

* How is life expectancy changing over time on different continents?

* Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy - you decide - a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

* Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged.

* Make up your own!


## Companion graphs

For each table, make sure to include a relevant figure.

Your figure does not have to depict every last number from the data aggregation result. Use your judgement. It just needs to complement the table, add context, and allow for some sanity checking both ways.

Notice which figures are easy/hard to make, which data formats make better inputs for plotting functions vs. for human-friendly tables.


## But I want to do more!

__Layout stretch goal:__ Get table and figure side-by-side. [This gist][jenny-plot-layout-gist] might get you started.

__Table stretch goal:__ There are some really nice fancy table helper packages. [This tweet from \@polesasunder][table-helper-pkgs-tweet] will point you toward some R packages you may want to check out (`pander`, `xtable`, `stargazer`).


## Report your process

You're encouraged to reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc. Give credit to your sources, whether it's a blog post, a fellow learner, an online tutorial, etc.


## Assess your work

The [general rubric](#lab-rubric) applies, but also...

__Check minus:__ Didn't tackle at least 3 tasks. Or didn't make companion graphs. Didn't interpret anything but left it all to the "reader". Or more than one technical problem that is relatively easy to fix. It’s hard to find the analysis or the narrative in all this R output.

__Check:__ Hits all the elements. No obvious mistakes. Pleasant to read. No heroic detective work required. Solid.

__Check plus:__ Exceeded the requirements in number of tasks. Or developed novel tasks that were indeed interesting and "worked". Impressive use of dplyr and/or ggplot2. Impeccable organization of repo and report. You learned something new from reviewing their work and you're eager to incorporate it into your work. The doc is very organized and well formatted.

[jenny-plot-layout-gist]: https://gist.github.com/jennybc/e9e9aba6ba18c72cec26
[table-helper-pkgs-tweet]: https://twitter.com/polesasunder/status/464132152347475968

```{r links, child="links.md"}
```