# Lab 02: Explore Gapminder and use `dplyr` {#lab-02}

<!--Original file: hw02_explore-gapminder-dplyr.md-->
<!--APH adaptation: https://ohsu-conj620.netlify.com/cm014.html-->

```{r include = FALSE}
source("common.R")
```

## Overview

The goal is to explore a new-to-you dataset. In particular, to begin to establish a workflow for data frames or “tibbles”. You will use `dplyr` and `ggplot2` to do some description and visualization.

Remember the [sampler][wiki-sampler] concept. Your work should serve as your own personal cheatsheet in the future for things to do with a new dataset. Give yourself the cheatsheet you deserve! Remember:

* `dplyr` should be your data manipulation tool
* `ggplot2` should be your visualization tool


## Bring rectangular data in

Work with the `gapminder` data. The Gapminder data is distributed as an [R package on CRAN][gapminder-cran].

Install `gapminder` if you have not done so already and remember to load it.

```{r eval = FALSE}
install.packages("gapminder")
library(gapminder)
```

Install and load `dplyr.` Probably via the `tidyverse` meta-package.

```{r eval = FALSE}
install.packages("tidyverse")
library(tidyverse)
```

## Smell test the data

Explore the `gapminder` object:

* Is it a data.frame, a matrix, a vector, a list?
* What's its class?
* How many variables/columns?
* How many rows/observations?
* Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?
* What flavor is each variable?
  
## Explore individual variables

Pick __at least__ one categorical variable and at least one quantitative variable to explore.

* Characterize what's possible, i.e. all possible values or max vs. min ... whatever's appropriate.
* What's typical? What's the spread? What's the distribution? Etc., tailored to the variable at hand.
* Feel free to use summary stats, tables, figures. We're NOT expecting high production value (yet).

## Explore various plot types

See the [`ggplot2` tutorial][ggplot2-tutorial], which also uses the `gapminder` data, for ideas. 

Make a few plots, probably of the same variable you chose to characterize numerically. Try to explore more than one plot type. **Just as an example** of what I mean:

* A scatterplot of two quantitative variables.
* A plot of one quantitative variable. Maybe a histogram or densityplot or frequency polygon.
* A plot of one quantitative variable and one categorical. Maybe boxplots for several continents or countries.

You don't have to use all the data in every plot! It's fine to filter down to one country or small handful of countries.

## Use `filter()`, `select()`, and `%>%`

Use `filter()` to create data subsets that you want to plot.

Practice piping together `filter()` and `select()`. Possibly even piping into `ggplot()`.


## But I want to do more!

*For people who want to take things further.*

Evaluate this code and describe the result. Presumably the analyst's intent was to get the data for Rwanda and Afghanistan. Did she succeed? Why or why not? If not, what is the correct way to do this?

```{r eval = FALSE}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

Read [What I do when I get a new data set as told through tweets][blog-post-new-data] from [SimplyStatistics][simply-statistics] to get some ideas!

Present numerical tables in a more attractive form, such as using `knitr::kable()`.

Use more of the `dplyr` functions for operating on a single table.

* [Introduction to dplyr][dplyr-intro]
* [`dplyr` functions for a single dataset][dplyr-single]
  
Adapt exercises from the chapters in the "Explore" section of [R for Data Science][r4ds] to the Gapminder dataset.

## Report your process

Reflect on what was hard/easy, problems you solved, helpful tutorials you read, etc. What things were hard, even though you saw them in class? What was easy(-ish) even though we haven't covered it yet?

## Assess your work

The [general rubric](#lab-rubric) applies, but also...

__Check minus:__ There are some mistakes or omissions, such as the number of rows or variables in the data frame. Or maybe no confirmation of its class or that of the variables inside. There are no plots or there's just one type of plot (and its probably a scatterplot). There's no use of `filter()` or `select()`. It’s hard to figure out where to look for what in this doc.

__Check:__ Hits all the elements. No obvious mistakes. Pleasant to read. No heroic detective work required. Solid.

__Check plus:__ Some "above and beyond", creativity, etc. You learned something new from reviewing their work and you're eager to incorporate it into your work now. Use of dplyr goes beyond `filter()` and `select()`. The `ggplot2` figures are quite diverse. The doc is very organized and well formatted.



[blog-post-new-data]: https://simplystatistics.org/2014/06/13/what-i-do-when-i-get-a-new-data-set-as-told-through-tweets/
[simply-statistics]: https://simplystatistics.org

```{r links, child="links.md"}
```